---
title: "Logistic Regression Modelling of the Probability of the First Marriage Ending in Divorce"
author: "Celine Kim (1001251147)"
date: "19 October 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Abstract

Using approximately 5000 observations on people's first marriage from the 2017
General Social Survey (GSS) on families, this report analyzes how the age of the
respondent and spouse at the start of the first marriage and the presence of
common-law unions influence the first marriage ending in divorce. The
multivariate logistic regression model used in this analysis revealed the
following relationship: 
<center>
$\log{\frac{p}{1-p}} = 1.6841 + 0.0648x_1 - 0.1136x_2 + 1.7476x_3$, 
</center>
where $p$ is the probability of the first marriage ending in
divorce, $x_1$ is the age of the respondent at the start of marriage, $x_2$ is
the age of the spouse or partner at the start of marriage, and $x_3$ is a binary
categorical variable that indicates the presence of a common-law union before
the first marriage. All p-values of the intercept and three explanatory
variables less than 0.05, standard errors less than 1, the result of Hosmer and
Lemeshow goodness of fit (GOF) test, and the Pearson's residual plot show that
this model is valid. However, since divorce is driven by multiple factors and
people who are currently in the first marriage have not been considered in this
report, a future study considering these factors will be needed for a more
accurate analysis.

## Introduction

The concept of common-law unions was very new to me as a person having lived in
South Korea for approximately 20 years. According to the Canadian Encyclopedia,
17.8% of families and 21.3% of all Canadian couples are from common-law unions.
Coming from a quite conservative cultural background where premarital
cohabitation is not that common, I was very interested in the different forms of
families in Canada and the relationship between marriage, common-law
relationships and divorce.
The goal of the analysis in this report is to find out how age of oneself and
partner at the start of one's first marriage and cohabitation experience from
common-law unions before their marriage influence the probability of divorce.
This paper will explore how the multivariate logistic regression model can be
used to analyze this relationship.

## Data

```{r, echo=FALSE, message=FALSE}
#install.packages("survey")
#install.packages("ResourceSelection")
library(tidyverse)
library(survey)
library(ggplot2)
library(dplyr)
library(ResourceSelection)

marriage_data <- read.csv("./gss_2017.csv")
```

The data used in this report is from the 2017 General Social Survey (GSS) on
families in Canada. This data provides information that increases understanding
of general social trends and characteristics of families that exist in Canada.
As the prime minister of Canada puts it, one of Canada's strengths is diversity,
which is evident in languages, cultural and ethnic backgrounds, gender and also
in families. This data was selected to explore patterns in the increasingly
diversifying forms and characteristics of families in Canada.

The survey asks questions that cover a wide range of topics. Some topics include
the respondent's date of birth, family origins, conjugal and parental history,
respondent's children and spouse or partner, organization within the household,
education, health, and much more. This report will focus on the section on
conjugal history of the respondents.

Data was collected from 2 February to 30 November 2017 through telephone
interviews with voluntary survey participation from the respondents. The
response rate of this survey was 52.4%.

As stated before, a strength of this questionnaire is the broad spectrum covered
about families. In particular, variables chosen for analysis in this report were
all able to be answered in numerical or categorial values that minimizes
subjectivity. In this questionnaire, an option called "valid skip" exists, which
checks for validity of the questions. For example, people on their first
marriage who have never experienced separation or divorce can choose the valid
skip option for questions regarding second marriage or divorce. Based on the
number of responses of valid skip for questions asking for the age of separation
or divorce questions and end type of first marriage, data seems to be reliable
because both the respondent and spouse categories are roughly 15,000. A drawback
of this survey is that there are too many questions in the questionnaire, which
usually influences the response rate and quality of responses. In addition,
another drawback is that some questions that ask for the respondent's age at the
time of parent's divorce or age of first marriage only covers from 15 to 49 and
any response over 50 is just recorded as 50 instead of the actual age.

Since this survey was conducted by collecting data at a specific point in time,
the 2017 GSS is a cross-sectional study. The target population is all people of
the age 15 and older in the 10 provinces of Canada. Residents of the Yukon,
Northwest Territories and Nunavut and full-time residents of institutions have
been excluded from the target population.

The survey frame was created by combining the list of valid landline and
cellular telephone numbers obtained by Statistics Canada and the Address
Register (AR) regardless if the phone number was linked to an address. However,
for those 86% of the phone numbers that were successfully connected to an
address on the AR, people with the same address  were grouped together using the
AR.

The 10 provinces were divided into 27 strata. The sampling technique used for
this survey is simple random sampling without replacement in each of the strata.
The respondents answered this survey through computer assisted telephone
interviews. The sample size for the 2017 GSS was 20,602, which was roughly
similar to the target sample size of 20,000.

Some trade-offs of this sampling approach is that people without phone numbers
have all been excluded. A drawback of the previously used random digit dialing
method was over-coverage; the survey was conducted without actually knowing
whether the random digits are a valid phone number so coverage can be greater
than the actual population. However, with this new method, since the list was
created at a certain point before the actual survey is conducted, service for
some of the telephone numbers might have been cancelled and new phone numbers
are not covered in the list which also creates over and under coverage issues.

The survey consists of questions that are required for weighting and those which
are not. Questions that are used for weighting do not allow a non-response
option. The calculated weights are then used to adjust for non-response. Also,
additional information available to Statistics Canada was used to adjust for
partial non-response. Finally, telephone numbers of non-responding interviewees
were dropped from the survey. The cost of this procedure is that since
non-responding numbers have been dropped, even if the weights are adjusted, the
sampling procedure does not strictly follow the simple random sampling method
considering that the response rate is around 50%.

## Model

The statistical model used in this analysis is the logistic regression model.
The software used to run the model is R. Given that the first marriage has
ended, we are modeling to see whether the first marriage will end because of
divorce or any other reasons. The aspects focused in this model are the age of
the respondent at the start of the first marriage, the age of the spouse at the
start of the first marriage, and whether the respondent and partner have engaged
in a common-law union before their marriage. Since end type of marriage is a
binary response variable where divorce is 1 and other reasons is 0, it was
thought that a logistic regression model was appropriate.

Let $x_1$ be the age of the respondent at the start of marriage. 

Let $x_2$ be the age of the spouse or partner at the start of marriage.

Let $x_3$ be a binary categorical variable where 1 represents that the
respondent was involved in a common-law union with the partner before the first
marriage and 0 represents otherwise.

The event of interest in this model is whether the first marriage will end
because of divorce.

Then let $p$ be the probability of the first marriage ending in divorce. 

Then this model in mathematical notation can be written as 
$\log{\frac{p}{1-p}} = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3$.

$\beta_0$ is the intercept which represents when $x_1$, $x_2$ and $x_3$ are 0.
However, in this specific model, it is not meaningful to interpret the intercept
because it is not possible for the respondent and partner to get marriage at age
0. $\beta_1$, $\beta_2$ and $\beta_3$ are coefficients that represent the
change in log odds for every one unit increase in $x_1$, $x_2$ and $x_3$
respectively.

In this model, $\log{\frac{p}{1-p}}$ is the log odds of the first marriage
ending in divorce. For example, if the probability of the first marriage ending
in divorce is 0.2, the odds will be $\frac{0.2}{1-0.2} = \frac{1}{4}$ which is a
4 to 1 odds against the first marriage ending in divorce.

The first and second aspect, the age of the respondent and partner at the start
of the first marriage, are important factors to consider when studying about
divorce because marriage at an early age is socially perceived as a red flag for
divorce. The Goldilocks Theory of Marriage states that the best age to get
married is between age 28 and 32. In addition, a sociologist at the University
of Utah named Nicholas Wolfinger, has found in a recent study that "divorce risk
declines for people who wait until their late 20s and early 30s to get married.
But it rises again for those who delay walking down the aisle until their late
30s". This shows that age at the start of one's marriage is an important aspect
that influences the risk of divorce.

According to the previously mentioned study, before the age of 32, the
probability of the marriage ending in divorce decreases by 11% each year. Since
a one year increase in age greatly influences the probability of divorce, the
numerical value of age was used in this model rather than categorical variables
such as age groups. In addition, since age was given in the data as numerical
values, such modelling decisions were possible.

The third aspect, which is common-law experience before marriage, is also an
important aspect in studying the risk of divorce. There is a common belief that
living together before marriage decreases the risk for divorce. However,
numerous studies including research conducted by Alan Booth and David Johnson
have concluded that premarital cohabitation actually leads to higher risk of
instability in marital life and divorce.

Since the study of interest deals with prescence of premarital cohabitation with
the partner instead of years of experience or age at the start of common-law
unions, a categorical variable was used in this model. For the following reason,
even though the age at the start of common-law relationships were given as
numerical data in the 2017 GSS data, such modelling decisions were found to be
appropriate.

According to the summary of this logistic regression model in Table 1 of the 
Appendix, R was able to estimate the odds ratio to find the best fit of this 
model. If the model does not converge, R fails to do so, which shows that
this model has converged. Also, the standard errors of the intercept and three
explanatory variables are less than 1, which indicates that there is no
overfitting problems (according to the Overfitting and Non-convergence lecture
from Imperial College London on coursera).

To check for the goodness of fit, the Hosmer-Lemeshow Test was used. A p-value
less than 0.05 indicates that the model is not a good fit. However, according to
the results of this model check in Table 2 of the Appendix, the p-value of 
0.3194 is greater than 0.05, which shows that this logistic regression model is 
a good fit.

```{r, echo=FALSE}
n = length(marriage_data$end_type_first_marriage)

# This is the 2017 population of Canada of age 15 and over calculated by using 
# information provided by Statistics Canada in a report called "Population 
# estimates on July 1st, by age and sex". 
N = 30633177

fpc.srs = rep(N,n)

marriage.design <- svydesign(id=~1, data=marriage_data, fpc=fpc.srs)

# Logistic regression using the Survey Library
mylogitsvy <- svyglm(end_type_first_marriage ~ age_at_first_marriage +
                       age_of_spouse_at_first_marriage +
                       as.factor(common_law_before_marriage), marriage.design, 
                     family="binomial")

# The Hosmer-Lemeshow Test to check for goodness of fit of the model
hl <- hoslem.test(mylogitsvy$y, fitted(mylogitsvy), g=10)
```

According to Figure 3 which shows the Pearson Residuals vs. Leverage plot 
in the result section (also found in the Appendix), since most points fall 
within the $\pm 3$ bandwidth without a specific pattern, we can say that the 
fitted logistic regression model is true.

A special case to be considered about this model is that since age of 50 and
above is recorded as 50 and age below 15 is outside the target population, this
model should not be used to extrapolate outside the data range of the
explanatory variables ($x_1, x_2 > 50$ or $x_1, x_2 < 15$).

An alternative model that can be used is a linear regression model if the
response variable is changed to the age of divorce of the first marriage. Also,
some variants that can be discussed are the presence of parent's divorce and the
age and existence of children from the first marriage.

## Results

Figure 1 is the scatterplot with the simple logistic regression line for the
explanatory variable, age of the respondent at the start of the first marriage,
and the response variable which is the end type of first marriage. Figure 2 is
the scatterplot with the same response variable but the explanatory variable is
age of the spouse at the start of the first marriage. These two plots show
opposite correlations according to the regression lines shown in the plots
below. Therefore, we will instead see how all three explanatory variables also
including the categorical variable affect the response variable by performing a
logistic regression analysis using the Survey Library.

```{r, echo=FALSE, message=FALSE, warning = FALSE}
ggplot(marriage_data, aes(x=age_at_first_marriage, y=end_type_first_marriage)) +
  geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) +
  ggtitle("Scatterplot of Age of Respondent at Start of First Marriage vs. \n 
          End Type of First Marriage")
```

##### Figure 1: Scatterplot of Age of Respondent at Start of First Marriage vs. End Type of First Marriage

<br />
<br />

```{r, echo=FALSE, message=FALSE, warning = FALSE}
ggplot(marriage_data, aes(x=age_of_spouse_at_first_marriage, y=end_type_first_marriage)) + 
  geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) +
  ggtitle("Scatterplot of Age of Spouse at Start of First Marriage vs. \n 
          End Type of First Marriage")
```

##### Figure 2: Scatterplot of Age of Spouse at Start of First Marriage vs. End Type of First Marriage

<br />

The summary of the logistic regression model using the Survey Library can be 
found in section Table 1 of the Appendix. The equation of the logistic 
regression model is 

$\log{\frac{p}{1-p}} = 1.6841 + 0.0648*$age_at_first_marriage 
$- 0.1136*$age_of_spouse_at_first_marriage
$+ 1.7476*$common_law_before_marriage, 

Now, I will show the equation where p is isolated. In this equation, the 
variable age_at_first_marriage was denoted as $x_1$, 
age_of_spouse_at_first_marriage is denoted as $x_2$ and 
common_law_before_marriage is denoted as $x_3$:

$p = \frac{e^{1.6841} \cdot e^{0.0648x_1} \cdot e^{1.7476x_3}}{(e^{0.1136x_2}+e^{1.6841} \cdot e^{0.0648x_1} \cdot e^{1.7476x_3})}$

<br />

Below is the Pearson Residuals vs. Leverage Plot used to check the model 
diagnostics.

```{r, echo=FALSE}
plot(mylogitsvy, which=5, main="The Pearson Residuals vs. Leverage Plot", 
     caption=c())
```

##### Figure 3: The Pearson Residuals vs. Leverage Plot

## Discussion

Out of the 20602 observations from the 2017 GSS data, a clean up procedure of
selecting the appropriate columns, changing unnecessary rows of data to "NA" and
extracting information needed for this analysis was first done. Then after
removing the "NA" data, 15650 rows were excluded. Therefore, 4952 observations
were used for this model, which is approximately 25% of the available data.

The equation of the model is 
$\log{\frac{p}{1-p}} = 1.6841 + 0.0648*$age_at_first_marriage 
$- 0.1136*$age_of_spouse_at_first_marriage
$+ 1.7476*$common_law_before_marriage. 

The intercept of 1.6841 represents when the explanatory variables
(age_at_first_marriage, age_of_spouse_at_first_marriage and
common_law_before_marriage) are 0. However, for this model, it is not meaningful
to interpret the intercept value because it is not possible for the respondent
and partner to get marriage at the age of 0. The coefficient value of 0.0648 of
age_at_first_marriage can be interpreted as the following: for every one year
increase in age, the log odds of the first marriage ending in divorce increases
by 0.0648. Secondly, the coefficient value of -0.1136 of
age_of_spouse_at_first_marriage can be interpreted as the following: for every
one year increase in age, the log odds of the first marriage ending in divorce
decreases by 0.1136. Finally, the categorical dummy variable
common_law_before_marriage is binary so it can either be 0 or 1. Thus, the
coefficient value of 1.7476 of common_law_before_marriage can be interpreted as
the following: the presence of a common law union between the respondent and
spouse before the first marriage increases the log odds of the first marriage
ending in divorce by 1.7476.

This shows that an increase in the age of the respondent at the start of the
first marriage has a positive correlation with the probability of the first
marriage ending in divorce while the age of the spouse or partner at the start
of the first marriage has a negative correlation with the probability of first
marriage ending in divorce. Among the explanatory variables considered in the
study, the existence of a common-law union between the respondent and spouse has
the greatest influence on the probability of divorce.

According to the summary of this logistic regression model in Table 1 of the
Appendix, p-values of the intercept and all three explanatory variables are less
than 0.05. This provides grounds to reject the null hypotheses that $\beta_0$,
$\beta_1$, $\beta_2$ and $\beta_3$ = 0.

While Table 3 of the Appendix shows the 95% confidence intervals of the log odds
ratio, Table 4 of the Appendix shows the 95% confidence intervals of the odds
ratio. In this part of the Discussion section, I will deal with the confidence
intervals of the odds ratio.

The 95% confidence interval of $\beta_0$, the intercept, is (3.8791845,
7.4827539).

The 95% confidence interval of age_at_first_marriage is (1.0486892, 1.0855979),
which means that we are 95% confident that a 1 year increase in the age of the
respondent at the start of the first marriage will result in an average increase
in probability of the first marriage ending in divorce as low as 1.0486892 or as
high as 1.0855979.

The 95% confidence interval of age_of_spouse_at_first_marriage is (0.8785619,
0.9068998), which means that we are 95% confident that a 1 year increase in the
age of the spouse at the start of the first marriage will result in an average
increase in probability of the first marriage ending in divorce as low as
0.8785619 or as high as 0.9068998.

The 95% confidence interval of the binary variable common_law_before marriage is
(4.7609321, 6.9216959), which means that we are 95% confident that the change
from 0 to 1 which indicates the presence of a common-law union between the
respondent and spouse before the first marriage in the age of the spouse at the
start of the first marriage will result in an average increase in probability of
the first marriage ending in divorce as low as 4.7609321 or as high as
6.9216959.

As discussed before in the Model Section, model convergence was shown by the
fact that R was able to estimate the odds ratio to find the best fit and
standard error values of less than 1. The p-value of 0.3194 in the result of the
Hosmer-Lemeshow Test which is greater than 0.05 showed that this logistic
regression model is a good fit. In addition, the Pearson Residuals vs. Leverage
plot showed that most points fall within the $\pm3$ bandwidth without a specific
pattern so it can be concluded that the fitted logistic regression model is
true.

One strange point is that the respondent's age shows a positive relationship
with the first marriage ending in divorce while the spouse's age moves in the
opposite direction. However, common-law unions are actually associated with
increased divorce rates as mentioned in the study done by Alan Booth and David
Johnson and its coefficient value of 1.747552 is greatest among all the
explanatory variables.

One caveat mentioned before about this logistic regression model is caution
about making inferences outside the data range. Since the questionnaire only
deals with respondents of age of 15 and older and records age of 50 and over as
50 instead of the actual age, we must not extrapolate outside the valid range of
15 to 50. Since there are numerous factors that are involved in divorce, this
model has some limitations in that it can show the general trend associated with
divorce, but might not be enough to explain divorce trends of the entire
Canadian population. Also, since this is just a probability model, it may not be
able to explain one specific case of divorce. Since this model considers the age
at the start of the first marriage without considering the current age of the
respondents or their spouses, it might not reflect the dramatic social changes
on the trend of marital status among the Canadian population during last several
decades. Actual numbers show that the general trend of marital status has
changed greatly in Canada. According to the Canadian Encyclopedia cited before,
while 92% of the families in Canada were married couples in 1961, there was a
decrease of 65.8% by the year 2016 and a drastic increase of common-law
relationships.

The next steps that I want to take in order to further develop this analysis is
to broaden the scope of subjects involved in the study. People who are currently
in their first marriage did not answer all four questions (age at the beginning
of first marriage, age of partner at the beginning of first marriage, common-law
experience before first marriage and end type of first marriage) in the
questionnaire. Instead, they only answered questions regarding their current
marriage. Therefore, in a future study, I would also like to involve those who
are currently in their first marriage.

In this future study, I would like to use the Bayesian statistical analysis
method. By combining prior information on the national average divorce rates
provided by Statistics Canada, I would like to make further statistical
inferences about the parameters of interest using analysis on the posterior
distribution. I would also like to conduct another survey that specifically
focuses on the factors that contribute to divorce to explore what explanatory
variables must be crucially involved to accurately model this relationship.


## References

Booth, A., & Johnson, D. (1988). Premarital Cohabitation and Marital Success.
Journal of Family Issues, 9(2), 255-272. doi:10.1177/019251388009002007

Bottle, A. (2019). Overfitting and Non-convergence - Assessing Model Fit.
Retrieved October 13, 2020, from
https://www.coursera.org/lecture/logistic-regression-r-public-health/overfitting-and-non-convergence-SUf7T

Common-Law Unions in Canada. (n.d.). Retrieved October 12, 2020, from
https://www.thecanadianencyclopedia.ca/en/article/common-law-unions-in-canada

Government of Canada, Statistics Canada. (2020, September 29). Population
estimates on July 1st, by age and sex. Retrieved October 15, 2020, from
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501

Statistics Canada. (2017). General social survey (GSS) on Family, 2017: Cycle 
31. [Public use microdata file and codebook]. Ottawa, ON: Statistics Canada. 
Retrieved from http://datacentre.chass.utoronto.ca/

The Pennsylvania State University. (2018). Model Diagnostics. Lecture. Retrieved
October 14, 2020, from https://online.stat.psu.edu/stat504/node/161/

Weissmann, J. (2015, July 16). Bad News: Waiting Too Long to Get Married May
Increase Your Risk of Divorce. Retrieved October 11, 2020, from
https://slate.com/business/2015/07/getting-married-late-increases-your-chance-of-a-divorce.html
