---
output: html_document
---
<h3> <center> Appendix </center> </h3>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Below is the packages and libraries used for this analysis. marriage_data is the
data extracted from the 2017 General Social Survey required for this study. 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
#install.packages("survey")
#install.packages("ResourceSelection")
library(tidyverse)
library(survey)
library(ggplot2)
library(dplyr)
library(ResourceSelection)

marriage_data <- read.csv("./gss_2017.csv")
```

### Table 1

Below is the summary of the logistic regression model using the Survey Library.
```{r}
n = length(marriage_data$end_type_first_marriage)

# This is the 2017 population of Canada of age 15 and over calculated by using 
# information provided by Statistics Canada in a report called "Population 
# estimates on July 1st, by age and sex". 
N = 30633177

fpc.srs = rep(N,n)

marriage.design <- svydesign(id=~1, data=marriage_data, fpc=fpc.srs)

# Logistic regression using the Survey Library
mylogitsvy <- svyglm(end_type_first_marriage ~ age_at_first_marriage +
                       age_of_spouse_at_first_marriage +
                       as.factor(common_law_before_marriage), marriage.design, 
                     family="binomial")
summary(mylogitsvy)
```

### Table 2

Below is the result of the Hosmer-Lemeshow Test to check for the goodness of fit 
of the model used in this analysis.
```{r}
# The Hosmer-Lemeshow Test to check for goodness of fit of the model
hl <- hoslem.test(mylogitsvy$y, fitted(mylogitsvy), g=10)
hl
```

### Table 3

Below is the 95% confidence interval calculated for log odds ratio.
```{r}
confint(mylogitsvy)
```

### Table 4

Below is the 95% confidence interval calculated for the odds ratio.
```{r}
exp(confint(mylogitsvy))
```

### Figure 1

```{r, message=FALSE, warning = FALSE}
ggplot(marriage_data, aes(x=age_at_first_marriage, y=end_type_first_marriage)) +
  geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) +
  ggtitle("Scatterplot of Age of Respondent at Start of First Marriage vs. \n 
          End Type of First Marriage")
```

\newpage

### Figure 2
```{r, message=FALSE, warning = FALSE}
ggplot(marriage_data, aes(x=age_of_spouse_at_first_marriage, y=end_type_first_marriage)) + 
  geom_point() + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) +
  ggtitle("Scatterplot of Age of Spouse at Start of First Marriage vs. \n 
          End Type of First Marriage")
```

\newpage

### Figure 3

```{r}
plot(mylogitsvy, which=5, main="The Pearson Residuals vs. Leverage Plot", 
     caption=c())
```
